Peptide retention time prediction

### 2015-10-08 response.tex

\breview

Other than that, there are some points worth mentioning. They will be
here presented in the order they appear on the manuscript.

On section 2, it is written ``so-called linear gradients''. Why say
``so-called''? This expression gives the idea that the gradients in
reality are not linear, but this isn’t further developed.

This is in part due to the strange nomenclature. In other fields a
gradient refers to the vector of spacial partial derivates, not a time
derivate.

However, to avoid confusion, we replaced the word so-called with
setting the words linear gradient in italics, to indicate that a
explanation of the term will follow.

@@

\ereview

We added a qualifying sentence to the caption of the Figure: ``It
should be noted that this relatively high reproducibility normally
only appears between runs under near to identical experiments, and do
not carry over into other experimental conditions or instrumentation.''

\breview

Across section 3, there is a lack of evaluation of the prediction
methods. In fact at a certain point it is said that one is ``good'',
one ``excellent'', and another that has ``lower'' accuracy followed by
one that is ``reasonable''. It is inferred that if these qualitative
comparisons are made by the authors, then there must be some accuracy
measure available. It would be preferable to have a table with these
values.

We removed all these unqualified references to methods accuracy.

@@

\breview

In section 3.2, it feels relevant to introduce the approach ELUDE takes regarding retention indexes. Even if they obtained via, are used as features for, a machine learning approach, it is still relevant to this section.
\ereview

We added a couple of sentences describing this.
\breview

Section 3.3 is rather brief. The reviewer feels there should be more information on these methods (or none at all).
\ereview

We added some more details on how BioLCCC works to the text.

On sections 3.1, 3.2 and 3.3 there is no evaluation of any of the
introduced methods. In section 3.4 we see the sort of qualitative
comparisons mentioned above. We note that there is no mention of how
the performance of ELUDE compares to the other methods, even if
qualitatively.

\ereview
We added a reference to a benchmark to the section.
\breview

If the proposed audience to this paper are scientists applying machine
learning methods to retention time prediction, more information on the
features used and the methods themselves would be relevant.

We included an brief enumeration of some of the features used in Elude.

@@

\breview

In the discussion the reviewer feels that it should be discussed not
only how retention time prediction can be improved, but also the
limitations of how accurate predictions can be, given the issues
mentioned above. Another relevant question is how much more relevant
do these predictions need be given the applications enumerated in
section 5.

\ereview

Estimate of how good predictions should be to be useful are generally
hard to make. Better performance will always be more useful than worse
performance.  Instead we added a comparison of the FPR of retention
time predictions versus predictions of iso-electric point calculations.

### 2015-10-08 retention.tex
answered all the questions

% SSRcalc was proposed including a number of additional terms; then it was extended
% for a few other column
Following these observations, Krokhin {\em et al.} developed SSRCalc
\cite{Krokhin2004}, which is currently the most widely used
index-based retention time predictor. In the initial version of the
algorithm a collection of 346 tryptic peptides was used to derive two
sets of retention coefficients, one corresponding to the N-terminal
and one to all the other positions in a peptide sequence. Furthermore,
correction factors for peptide length and overall hydrophobicity were
included. Subsequent versions of the algorithm included terms
accounting for additional properties such as isoelectric point,
nearest-neighbor effects of charged peptides and propensity to form
helical structures \cite{Krokhin2006}. Recently, it was showed that
interactions involving helical structures are highly complex, and thus
more sophisticated models are needed to account for the effects that
such interactions have on peptide retention time \cite{ah}.

@@

In other modeling-based approaches theoretical models are used to
capture the chemical behavior of peptides on the chromatography
column. One such method is BioLCCC, which attempts to model the
connectivity of the amino acids in a peptide chain and the general
mechanisms of chromatographic
separation \cite{gorshkov2006}. Nevertheless, given the numerous
factors involved in peptide separation, modeling approaches often
involve a considerable number of simplifications.

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector
machines (SVMs) \cite{klammer2007improving}. Their method required
fewer training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction accuracy.
Subsequently, a more elaborate SVM-based method using a customized
kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
approach provided an algorithm that required as little as 50 peptides
for training.

@@

Moruz {\em et al.} \cite{elude1} developed a software package for
retention time prediction named {\sc Elude}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. This ensured that the algorithm
could be applied to different chromatography setups with minimal loss
in performance. For cases when little training data was available,
{\sc Elude} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

However, despite all these efforts, it should be pointed out that none
of the retention time prediction methods currently available is able
to provide a prediction precision comparable to the small shifts
observed when comparing retention time of peptides from two technical
replicates (Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further studies are
required to even understand the error sources in the process.

>>>>

% SSRcalc was proposed including a number of additional terms; then it was extended
% for a few other column
Following these observations, Krokhin {\em et al.} developed {\sc
SSRCalc} \cite{Krokhin2004}, which is currently the most widely used
index-based retention time predictor. In the initial version of the
algorithm a collection of 346 tryptic peptides was used to derive two
sets of retention coefficients, one corresponding to the N-terminal
and one to all the other positions in a peptide sequence. Furthermore,
correction factors for peptide length and overall hydrophobicity were
included. Subsequent versions of the algorithm included terms
accounting for additional properties such as isoelectric point,
nearest-neighbor effects of charged peptides and propensity to form
helical structures \cite{Krokhin2006}. Recently, it was showed that
interactions involving helical structures are highly complex, and thus
more sophisticated models are needed to account for the effects that
such interactions have on peptide retention time \cite{ah}.

@@

In other modeling-based approaches theoretical models are used to
capture the chemical behavior of peptides on the chromatography
column. One such method is BioLCCC, models the chromatography process
using a lattice based random walk model for the peptides, which are
modeled as chains of monomers. The model also includes an entropy and
energy compensation for the amount of peptides residing on the column 
and a model of the interaction energies between the peptide and the
stationary phase.

Nevertheless, given the numerous factors involved in peptide
separation, modeling approaches often involve a considerable number of
simplifications.

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector
machines (SVMs) \cite{klammer2007improving}. Their method required
fewer training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction accuracy.
Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
 use a customized kernel function was proposed \cite{rtpredict,
rtpredictImproved}. This approach provided an algorithm that required
as little as 50 peptides for training.

@@

Moruz {\em et al.}~\cite{elude1} developed a software package for
retention time prediction named {\sc Elude}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. The features included properties
such as average hydrophobicity, the N- and C-termini's hydrophobicity,
the stretch of the five most and least hydrophobic amino acids,
helicity, and the number of occurrences of the 25\% most or least
hydrophobic amino acids. This ensured that the algorithm could be
applied to different chromatography setups with minimal loss in
performance. For cases when little training data was available, {\sc
Elude} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

In a benchmark based on the merit of how large retention time window,
measured as a fraction of the gradient time was be needed to include
95\% of the peptides of a series of 14 runs, {\sc Elude} outperformed
{\sc SSRCalc} and {\sc RTPredict}, needing only on average 22\% as
compared to 28\% and 26\% respectively~\cite{elude1}. However, despite
all these efforts, it should be pointed out that none of the retention
time prediction methods currently available is able to provide a
prediction precision comparable to the small shifts observed when
comparing retention time of peptides from two technical replicates
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further studies are
required to even understand the error sources in the process.
